---
title: "Colour Analysis - Phenotypic Plasticity at the Initiation of Dominance hierarchies"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: hide
---
<style>
  h2{color: #DC7633 !important}
  h1{color: #5499C7 !important}
  body{background-color: white !important}
</style>

<style>
.list-group-item.active, .list-group-item.active:hover, .list-group-item.active:focus {
    color: #212F3D;
    background-color: #EDBB99;
    font-weight: bold;}
a {
    color: #5499C7;}
.nav-tabs > li.active > a, .nav-tabs > li.active > a:hover, .nav-tabs > li.active > a:focus {
    color: #DC7633;}
body {
    font-family: "Open Sans","Helvetica Neue",Helvetica,Arial,sans-serif;
    font-size: 15px;
    line-height: 1.42857143;
    color: #212F3D;}
</style>


<br/>

```{r set some defaults, echo = FALSE}

library("knitr")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
pdf.options(useDingbats = FALSE)
options(knitr.kable.NA = '')

```


```{r load packaged, echo=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(multcomp)
library(tidyr)
library(lmerTest)
library(emmeans)
library(interactions)
library(kableExtra)
library(MuMIn)
library(grid)
```

## Introduction

https://www.rapidtables.com/web/color/RGB_Color.html - website to get hex codes from hue/RGB values



```{r, include=FALSE}
all_colour_data <- read.csv("../Data/Vizer_Strategic_growth_colour_data_SNP_corrected_20250506.csv", header = T)
head(all_colour_data)


colour_subset_data <- subset(all_colour_data, !replicate_ID == "8" & !replicate_ID == "9" & !replicate_ID == "18")
head(colour_subset_data)
colour_subset_data$replicate_ID #all good
table(colour_subset_data$fish_type2) #all good

#extract columns of interest:
colour_subset_data <- colour_subset_data[ , c("replicate_ID", "housing_tank_ID", "fish_type2",  "initial_hue", "final_hue", "initial_saturation", "final_saturation")]
head(colour_subset_data)
```

### List of replicates to use
```{r}
print(colour_subset_data$replicate_ID)
table(colour_subset_data$fish_type2)
```


```{r}
head(colour_subset_data)

#extract columns of hue:
colour_subset_data_hue <- colour_subset_data[ , c("replicate_ID", "housing_tank_ID", "fish_type2",  "initial_hue", "final_hue")]
head(colour_subset_data_hue)

hue_data <- gather(colour_subset_data_hue, key = "time", value = "hue", 4:5)
hue_data

#extract columns of saturation:
colour_subset_data_satu <- colour_subset_data[ , c("replicate_ID", "housing_tank_ID", "fish_type2","initial_saturation", "final_saturation")]
head(colour_subset_data_satu)

satu_data <- gather(colour_subset_data_satu, key = "time", value = "satu", 4:5)
satu_data
```

## Linear-mixed model {.tabset}

### Hue
Hue over time across three fish categories
$formula: hue$ ~ $time*FishType2 + (1|replicateID)$
```{r}
hue_data$time[hue_data$time == "initial_hue"] <- "week 0"
hue_data$time[hue_data$time == "final_hue"] <- "week 5"
head(hue_data)

#time needs to be a factor
hue_data$time <- as.factor(hue_data$time)

#fish_type2 needs to be a factor
hue_data$fish_type2 <- as.factor(hue_data$fish_type2)

#replicateID needs to be a factor
hue_data$replicate_ID <- as.factor(hue_data$replicate_ID)
```


```{r}
mixed_mod_hue <- lmer(hue ~ time*fish_type2 + (1|replicate_ID), data = hue_data)
anova(mixed_mod_hue)
```

interaction not significant:

```{r}
mixed_mod_hue_s <- step(mixed_mod_hue,reduce.fixed = TRUE, reduce.random = TRUE)
plot(mixed_mod_hue_s)
mixed_mod_hue_s
```

run best model:
```{r}
mixed_mod_hue_final <- lmer(hue ~ time+fish_type2 + (1|replicate_ID), data = hue_data)
anova(mixed_mod_hue_final)
```
R-squared values of the best fit model:
```{r}
r.squaredGLMM(mixed_mod_hue_final)
```

### Post-Hoc comparisons
```{r}
emm <- emmeans(mixed_mod_hue_final, ~ time + fish_type2)

emms_tbl <- pairs(emm, simple = "each") %>%
  rbind() 

emms_tbl%>%
   kbl(col.names = c("Fish Category", "Time", "Contrast", "Estimate", "SE", "df", "t ratio", "p-value"), caption = "Simultaneous pair-wise post-hoc comaprisons of size of fish categories (P1, P2, S) and time (week 0, week 5).", align = "c", digits = 3) %>%
  kable_classic(full_width = F, html_font = "Cambria")%>%
  column_spec(1:3, bold = T) %>%
  row_spec(0, color = "white", background = "black",bold = T)

```


### Figures
```{r summary statistics, include=FALSE}
hue_data_summary <- ddply(hue_data, c("time", "fish_type2"),summarise,
               N    = sum(!is.na(hue)),
               mean = mean(hue, na.rm=TRUE),
               sd   = sd(hue, na.rm=TRUE),
               se   = sd / sqrt(N) )
hue_data_summary
```

```{r setting up data for figure, include=FALSE}
head(hue_data)
hue_data_summary
colnames(hue_data_summary)[4] = "hue"
```


```{r size figure over time, fig.height=7, fig.width=7}
gg_hue <- ggplot(hue_data, aes(x=time, y=hue,group=fish_type2, colour=fish_type2))+
        geom_jitter(position = position_jitterdodge(0.8), size=4, alpha=0.3) +
        geom_line(aes(group=fish_type2), data = hue_data_summary,  position=position_dodge(0.8), size=1)+
        geom_pointrange(aes(ymin = hue-se, ymax = hue+se), data =hue_data_summary, size=1.2, alpha=1, position=position_dodge(0.8))+
        scale_color_manual(values = c( "magenta3","purple3", "grey50"), name ="social poisition") +
        theme_classic() +
        theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position = c(0.15,0.9),         legend.title=element_text(size=16),  legend.text=element_text(size=16))+
        scale_y_continuous(trans = "reverse") +
        ylab("Hue value + SEM") + 
        xlab("Pre-experiment Post-experiment")
gg_hue

#ggsave("reverse_y_hue.png", height = 7, width = 7, dpi = 1200)

```



### Saturation
Saturation over time across three fish categories
$formula: saturation$ ~ $time*FishType2 + (1|replicateID)$
```{r}
satu_data$time[satu_data$time == "initial_saturation"] <- "week 0"
satu_data$time[satu_data$time == "final_saturation"] <- "week 5"
head(satu_data)

#time needs to be a factor
satu_data$time <- as.factor(satu_data$time)

#fish_type2 needs to be a factor
satu_data$fish_type2 <- as.factor(satu_data$fish_type2)

#replicateID needs to be a factor
satu_data$replicate_ID <- as.factor(satu_data$replicate_ID)
```


```{r}
mixed_mod_satu <- lmer(satu ~ time*fish_type2 + (1|replicate_ID), data = satu_data)
anova(mixed_mod_satu)
```


```{r}
mixed_mod_satu_s <- step(mixed_mod_satu,reduce.fixed = TRUE, reduce.random = TRUE)
plot(mixed_mod_satu_s)
mixed_mod_satu_s
```
run best model:
```{r}
mixed_mod_satu_final <- lm(satu ~ time, data = satu_data)
anova(mixed_mod_satu_final)
```

### Figures
```{r summary statistics, include=FALSE}
satu_data_summary <- ddply(satu_data, c("time", "fish_type2"),summarise,
               N    = sum(!is.na(satu)),
               mean = mean(satu, na.rm=TRUE),
               sd   = sd(satu, na.rm=TRUE),
               se   = sd / sqrt(N) )
satu_data_summary
```

```{r setting up data for figure, include=FALSE}
head(satu_data)
satu_data_summary
colnames(satu_data_summary)[4] = "satu"
```


```{r size figure over time, fig.height=7, fig.width=7}
gg_satu <- ggplot(satu_data, aes(x=time, y=satu,group=fish_type2, colour=fish_type2))+
        geom_jitter(position = position_jitterdodge(0.8), size=4, alpha=0.3) +
        geom_line(aes(group=fish_type2), data = satu_data_summary,  position=position_dodge(0.8), size=1)+
        geom_pointrange(aes(ymin = satu-se, ymax = satu+se), data =satu_data_summary, size=1.2, alpha=1, position=position_dodge(0.8))+
        scale_color_manual(values = c( "magenta3","purple3", "grey50"), name ="social position") +
        theme_classic() +
        theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position = "righ",         legend.title=element_text(size=16),  legend.text=element_text(size=16))+
        ylab("Saturation value + SEM") + 
        xlab("Pre-experiment Post-experiment")
gg_satu

#ggsave("saturation.png", height = 7, width = 7, dpi = 1200)

```

```{r}
#combine figures
#gg_hue + gg_satu
library(cowplot)

gg_hue <- ggplot(hue_data, aes(x=time, y=hue,group=fish_type2, colour=fish_type2))+
        geom_jitter(position = position_jitterdodge(0.8), size=4, alpha=0.3) +
        geom_line(aes(group=fish_type2), data = hue_data_summary,  position=position_dodge(0.8), size=1)+
        geom_pointrange(aes(ymin = hue-se, ymax = hue+se), data =hue_data_summary, size=1.2, alpha=1, position=position_dodge(0.8))+
        scale_color_manual(values = c( "magenta3","purple3", "grey50"), name ="social poisition") +
        theme_classic() +
        theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position = c(.85,0.85),        
              legend.title=element_text(size=16),  legend.text=element_text(size=16))+
        #scale_y_continuous(trans = "reverse") +
        ylab("Hue value + SEM") + 
        xlab("Experimental week") 
gg_hue


gg_satu <- ggplot(satu_data, aes(x=time, y=satu,group=fish_type2, colour=fish_type2))+
        geom_jitter(position = position_jitterdodge(0.8), size=4, alpha=0.3) +
        geom_line(aes(group=fish_type2), data = satu_data_summary,  position=position_dodge(0.8), size=1)+
        geom_pointrange(aes(ymin = satu-se, ymax = satu+se), data =satu_data_summary, size=1.2, alpha=1, position=position_dodge(0.8))+
        scale_color_manual(values = c( "magenta3","purple3", "grey50"), name ="social position") +
        theme_classic() +
        theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position = "righ",         legend.title=element_text(size=16),  legend.text=element_text(size=16))+
        ylab("Saturation value + SEM") + 
        xlab("Experimental week")
gg_satu



combined_plot <- plot_grid(gg_hue, gg_satu, labels = c("A", "B"), ncol = 2, align = "v", rel_heights = c(1, 1))
print(combined_plot)

#save
ggsave("../Figures/combined_hue_saturation.png", plot = combined_plot, width = 14, height = 7, dpi = 1200, bg="white")

#save pdf
#ggsave("combined_hue_saturation.pdf", plot = combined_plot, width = 14, height = 7, dpi = 1200)


```



#### Figures with color background:

```{r}
#hue:
reds <- c("#FF670D","#FFA640")
g <- rasterGrob(scales::alpha(reds, 1), width=unit(1,"npc"), height = unit(1,"npc"), 
interpolate = TRUE) 

gg_hue_color <- ggplot(hue_data, aes(x=time, y=hue,group=fish_type2, shape=fish_type2)) + # add gradient background 
   annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) + 
        geom_jitter(position = position_jitterdodge(0.3), size=4, alpha=0.6) +
        geom_line(aes(group=fish_type2), data = hue_data_summary,  position=position_dodge(0.8), size=1, color="grey80")+
        geom_pointrange(aes(ymin = hue-se, ymax = hue+se), data =hue_data_summary, size=1.2, lwd=1.2, alpha=1, position=position_dodge(0.8), color="grey80")+
        theme_classic() +
        theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position ="right", legend.title=element_text(size=16),  legend.text=element_text(size=16))+
        scale_y_continuous(trans = "reverse") +
    labs(shape = "fish category")+
        ylab("Hue value + SEM") + 
        xlab("Pre-experiment Post-experiment") # add data layer  
gg_hue_color

ggsave("../Figures/hue_with_color_bg.png", height = 7, width = 7, dpi = 1200)
```


```{r}
#saturation

gradient <- c("#FF850D","#FFD8B3")
g1 <- rasterGrob(scales::alpha(gradient, 1), width=unit(1,"npc"), height = unit(1,"npc"), 
interpolate = TRUE) 

gg_satu_color <- ggplot(satu_data, aes(x=time, y=satu,group=fish_type2, shape=fish_type2))+
      annotation_custom(g1, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) + 
        geom_jitter(position = position_jitterdodge(0.3), size=4, alpha=0.6) +
        geom_line(aes(group=fish_type2), data = satu_data_summary,  position=position_dodge(0.8), size=1, color="grey80")+
        geom_pointrange(aes(ymin = satu-se, ymax = satu+se), data =satu_data_summary, size=1.2, alpha=1, position=position_dodge(0.8), color="grey80")+
        theme_classic() +
        theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position ="righ",         legend.title=element_text(size=16),  legend.text=element_text(size=16))+
        ylab("Saturation value + SEM") + 
        xlab("Pre-experiment Post-experiment")

gg_satu_color

#ggsave("saturation_with_color_bg.png", height = 7, width = 7, dpi = 1200)
```



