---
title: "Phenotypic_Plasticity_gene_expression"
author: "Vizer Lili"
date: "2023-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load packages, include=FALSE}

library(tidyverse)
library(dplyr)
#library(arrayQualityMetrics) # need special install above
library(ggplot2)
library(readr)
library(plotly)
library(DESeq2) # need special install above
library(vegan)
library(data.table)
library(ggpubr)
library(pdftools)
library(ggrepel)
library(adegenet)
library(ggvenn)
library(flashClust)
library(WGCNA)
library(gridExtra)
###conduct array quality metrics to detect and remove outliers
#library(affycoretools)
#library(genefilter)
library(Biobase)
library(gridExtra)

source("CustomFunctions.R")

#color scheme throughout
#write out color-scheme
color_scheme <- c("#abd9e9", "#2c7bb6", "darkblue", "#fdae61", "darkorange",  "#d7191c", "pink1", "magenta", "purple") 

col_vector<-c('#e6194b', '#3cb44b', '#ffe119', '#4363d8', '#f58231', '#911eb4', '#46f0f0', '#f032e6', '#bcf60c', '#fabebe', '#008080', '#e6beff', '#9a6324', '#fffac8', '#800000', '#aaffc3', '#808000', '#ffd8b1', '#000075', '#808080', "green3", '#000000', "#abd9e9",  "#d7191c", "#2c7bb6","maroon", "green", "yellow", "coral" , "gray85")
```


## Expoloring data {.tabset}

```{r}
# read in the counts file
counts <- read.table("../Data/nemo_featurecounts_24July2023_.txt", header = TRUE, row.names = 1)
counts <- counts[c(-1:-5)] # removing columns 1-5 since we do not need them for this
head(counts)
col_names <- colnames(counts)

# Remove some of the extra stuff in the column names to match with expDesign
col_names <- gsub("X", "", col_names)
col_names <- gsub("\\.*", "", col_names)
colnames(counts) <- col_names
head(counts)

#Need to re-arrange count data so the first columns are S, followed by P1 and P2 - S will be used as control here
reorder_dataframe <- function(counts, custom_order) {
  col_order <- order(match(sub("[0-9]+([A-Za-z])", "\\1", colnames(counts)), custom_order))
  df_reordered <- counts[, col_order, drop = FALSE]  # Use drop = FALSE to keep as a dataframe
  return(df_reordered)
}

# Specify the custom order you want
desired_order <- c("S", "P1", "P2")

# Reorder the dataframe
counts <- reorder_dataframe(counts, desired_order)

#Write clean counts file (eg.: pathway-analysis)
head(counts)
#write.table(counts, file = "../Data/nemo_cleaned_count_data.txt", sep = "\t", row.names = TRUE, col.names = TRUE)

# read in the experimental design .csv
expDesign <- read.csv("../Data/Strategic_Growth_Exp_Design.csv")
head(expDesign)
expDesign$fish_type2 <- factor(expDesign$fish_type2)
expDesign$fish_type <- factor(expDesign$fish_type)
expDesign$clutch_ID <- factor(expDesign$clutch_ID)
expDesign$Sample_ID <- factor(expDesign$Sample_ID)

table(expDesign$fish_type2, expDesign$clutch_ID)
```

```{r, raw counts}
totalCounts=colSums(counts)
totalCounts
barplot((totalCounts), ylab="raw counts", las=2, cex.names=0.5, cex.axis = 0.8)
```

<br/>

## conduct array quality metrics to detect and remove outliers {.tabset}

```{r array of quality metrics to detect and remove outliers}

## create the DESeq object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = expDesign,
                              design = ~ clutch_ID + fish_type2)

vsd.ge=assay(vst(dds))
rl=vst(dds)
#e=ExpressionSet(assay(rl), AnnotatedDataFrame(as.data.frame(colData(rl))))
#arrayQualityMetrics(e,outdir=v,intgroup=c("clutch_ID", "fish_type2"),force=T)
```



<br/>

### Filtering counts

```{r unfiltered size factor plots, fig.width = 10, fig.height = 4}

nrow <- nrow(counts) # number of rows/counts (24840)
countMat <- DESeqDataSetFromMatrix(counts, expDesign, ~ clutch_ID + fish_type2) # makes a DESeqDataSet object with count data, experimental design, and formula: ~ clutch_ID + fish_type2
counts_SF <- estimateSizeFactors(countMat) #  estimates the size factors using the "median ratio method" described by Equation 5 in Anders and Huber (2010)

# make dataframe of the size factors to visualize
sizeFactors <- data.frame(sample = counts_SF@colData@listData[["Sample_ID"]], treat = counts_SF@colData@listData[["fish_type2"]], sizeFactors = counts_SF@colData@listData[["sizeFactor"]])

# plot sizeFactors
ggplot(data = sizeFactors, aes(x = sample, y = sizeFactors, fill = treat)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  #scale_fill_brewer(palette = 2, type = "qual") +
  geom_bar(stat = "identity") +
  ggtitle("Unfiltered sizeFactors")
```

Figure 1: Size factors of initial clownfish samples in the study, calculated using *A. percula* gene count data.

After viewing size factors, it appears that no samples need to be removed do to low values.

</br>


```{r filter low reads (base mean 3)}

### Filtered counts and remove outliers (less than 10)

## remove rows with no or low counts (remove ones with base mean 10 or lower)
keep <- rowSums(counts(counts_SF)) >= 10 
counts_filter <- counts[keep,]
nrow_filter <- nrow(counts_filter)
nrow_filter
# the updated number of rows (20537)  #with the above change of 10 this is 18829
filter_counts_out <- counts_filter[]

```


```{r filtered size factor plots, fig.width = 10, fig.height = 4}

## Redo the count matrix steps from above with the filtered data
filt_countMat <- DESeqDataSetFromMatrix(filter_counts_out, expDesign, ~ clutch_ID + fish_type2) # makes a DESeqDataSet object with count data, experimental design, and design formula of ~fishtype_2 + clutch_ID

counts_SF_filter <- estimateSizeFactors(filt_countMat) #  estimates the size factors using the "median ratio method" described by Equation 5 in Anders and Huber (2010)


# make dataframe of the size factors to visualize
sizeFactors_filter <-  data.frame(sample = counts_SF_filter@colData@listData[["Sample_ID"]], treat = counts_SF_filter@colData@listData[["fish_type2"]], sizeFactors = counts_SF_filter@colData@listData[["sizeFactor"]])

# plot sizeFactors
ggplot(data = sizeFactors_filter, aes(x = sample, y = sizeFactors, fill = treat)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_bar(stat = "identity") +
  ggtitle("Filtered sizeFactors")

```
Figure 2: Size factors of clownfish samples after filtering low reads out.

Without filtering low reads, we had a total of **`r nrow`** counts. After filtering out the low counts (those with a base mean less than 3), we now have **`r  nrow_filter`** counts remaining for all *A. percula* samples.



```{r dds model and save, eval = FALSE, echo = TRUE}

# going to use the filter_counts_out object created above because it is filtered for low reads 
## create the DESeq object
dds <- DESeqDataSetFromMatrix(countData = filter_counts_out,
                              colData = expDesign,
                              design = ~ clutch_ID + fish_type2)
dds <- DESeq(dds) # differential expression analysis on gamma-poisson distribution
vsd <- varianceStabilizingTransformation(dds, blind = TRUE) # quickly estimate dispersion trend and apply a variance stabilizing transformation

## saving the rlog for DEG heatmap
rlog <- DESeq2::rlog(dds, blind = TRUE) #for use later on

## Save dds and vsd data 
save(dds, vsd, rlog, file = "./transformed_counts_mindepth10.RData")

#Look at dispersions plot
plotDispEsts(dds, main="Dispersion plot")

```


```{r normalized count table}
## Load previously saved dds, vsd, and rlog data 
load("transformed_counts_mindepth10.RData")

## Normalize counts for table
dds <- estimateSizeFactors(dds)
norm_counts <- counts(dds, normalized = TRUE)

```

### PCA plots of Global Gene Expression 

```{r}
## PCA plot using DESeq2 package of all genes vs vegan package
## Principal Component Analysis
PCAdata_all <- plotPCA(vsd, intgroup=c("clutch_ID", "fish_type2"), ntop=NULL, returnData=TRUE)
percentVar_all <- round(100 * attr(PCAdata_all, "percentVar"))
PCAdata_all

## Making plot - DESeq2 pca
color_scheme <- c("magenta3","purple3","grey50")
pca_plot_all <- ggplot(PCAdata_all, aes(x = PC1, y = PC2, color = fish_type2)) +
  geom_point(size = 4, aes(fill = fish_type2, shape=clutch_ID), alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2, fill = fish_type2), fill = NA, type = "t", level = 0.95, alpha = 0.2, show.legend = FALSE) +
#  geom_text(aes(label = name), vjust = -0.5, hjust = 0.5, size = 2, color = "black") +  # Add this line to include sample names as labels+
  xlab(paste0("PC1 (", percentVar_all[1], "%)")) +
  scale_fill_manual(name = "Treatment", values=color_scheme) +
  scale_color_manual(name = "Treatment", values=color_scheme) +
  ylab(paste0("PC2 (", percentVar_all[2], "%)")) +
  ggtitle("All genes") +
  theme_classic()
pca_plot_all

```


```{r}
## PCA using prcomp from vegan package
## Extract variance-stabilized data from DESeq2
pca_data_all <- assay(vsd)  # or use rlog(dds) if preferred

# Transpose for PCA (samples as rows, genes as columns)
pca_res <- prcomp(t(pca_data_all), center = TRUE, scale. = FALSE)

# Create a data frame with PCA results
pca_df <- as.data.frame(pca_res$x)
pca_df$fish_type2 <- colData(vsd)$fish_type2
pca_df$clutch_ID <- colData(vsd)$clutch_ID

# Compute variance explained by each PC
percentVar <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

color_scheme <- c("magenta3", "purple3", "grey50")
shapes <- c(15,  16,  17)



## Plot PCA (PCA1 vs PCA2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = fish_type2)) +
  geom_point(size = 4, aes(shape = clutch_ID), alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2), 
               fill = NA, type = "t", level = 0.95, alpha = 0.2, show.legend = FALSE) +
  xlab(paste0("PC1 (", percentVar[1], "%)")) +
  ylab(paste0("PC2 (", percentVar[2], "%)")) +
  scale_color_manual(name = "social position", values = color_scheme) +
  ggtitle("PCA (PC1 vs PC2) all genes prcomp") +
  theme_classic()

#ggsave("../Figures/PCA_PC1_vs_PC2_all_genes_prcomp.png", width = 10, height = 8, dpi = 1200)

## Plot PCA (PC1 vs PC2) - manuscript version (supplements)
pca_plot12 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = fish_type2)) +
  geom_point(size = 5, aes(shape = clutch_ID), alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2), 
               fill = NA, type = "t", level = 0.95, alpha = 0.2, size = 2,
               show.legend = FALSE) +
  xlab(paste0("PC1 (", percentVar[1], "%)")) +
  ylab(paste0("PC2 (", percentVar[2], "%)")) +
  scale_color_manual(name = "social position", values = color_scheme) +
  scale_shape_manual(name = "clutch ID", values = shapes) +
  annotate("text", x = Inf, y = -Inf,                    #bottom right: Inf, y = -Inf, hjust = 1.1, vjust = -0.5
           label = "ADONIS p(social position) = 0.004",
           hjust = 1.1, vjust = -0.5, 
           size = 4, fontface = "italic") +
  #ggtitle("PCA (PC1 vs PC2) all genes prcomp") +
  theme_classic() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = c(0.09, 0.18))
#ggsave("../Figures/PCA_PC1_vs_PC2_all_genes_prcomp.pdf", width = 10, height = 8, dpi = 1200)
#ggsave("../Figures/PCA_PC1_vs_PC2_all_genes_prcomp.png", width = 10, height = 8, dpi = 1200)


## PERMANOVA test on PCA 1 and PCA 2
pca_matrix_pc1_pc2 <- pca_df[, c("PC1", "PC2")]
adonis_res_pc1_pc2 <- adonis2(pca_matrix_pc1_pc2 ~ fish_type2 + clutch_ID, 
                              data = pca_df, 
                              method = "euclidean", 
                              permutations = 999)
print(adonis_res_pc1_pc2)

#Permutation test for adonis under reduced model
#Terms added sequentially (first to last)
#Permutation: free
#Number of permutations: 999

#adonis2(formula = pca_matrix_pc1_pc2 ~ fish_type2 + clutch_ID, data = pca_df, permutations = 999, method = "euclidean")
#           Df SumOfSqs      R2      F Pr(>F)   
#fish_type2  2   4305.4 0.15859 4.1744  0.004 **
#clutch_ID   2   2214.1 0.08156 2.1467  0.083 . 
#Residual   40  20627.8 0.75985                 
#Total      44  27147.2 1.00000                 
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1



```




```{r #PCA - PC2&PC3 - main text PCA plot}

p_val <- adonis_res["fish_type2", "Pr(>F)"]
p_label <- if (is.na(p_val)) {
  "ADONIS p(social position) = NA"
} else if (p_val < 0.001) {
  "ADONIS p(social position) < 0.001"
} else {
  sprintf("ADONIS p(social position) = %.3f", p_val)
}

# Plot PCA (PC2 vs PC3)
pca_plot <- ggplot(pca_df, aes(x = PC2, y = PC3, color = fish_type2)) +
  geom_point(size = 5, aes(shape = clutch_ID), alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2), 
               fill = NA, type = "t", level = 0.95, alpha = 0.2, size= 2, 
               show.legend = FALSE) +
  xlab(paste0("PC2 (", percentVar[2], "%)")) +
  ylab(paste0("PC3 (", percentVar[3], "%)")) +
  scale_color_manual(name = "social position", values = color_scheme) +
  scale_shape_manual(name = "clutch ID", values = shapes) +  # Specify shape values
  #ggtitle("PCA (PC2 vs PC3) all genes prcomp") +
  theme_classic() +
  theme(axis.text = element_text(size = 16), axis.title = element_text(size = 16),         legend.title=element_text(size=13),  legend.text=element_text(size=13), legend.position = c(0.09,0.18))
#ggsave("../Figures/PCA_PC2_vs_PC3_all_genes_prcomp.png", width = 10, height = 8, dpi = 1200)

# Plot PCA (PC2 vs PC3) - manuscript version
pca_plot <- ggplot(pca_df, aes(x = PC2, y = PC3, color = fish_type2)) +
  geom_point(size = 5, alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2), 
               fill = NA, type = "t", level = 0.95, alpha = 0.2, size= 2, 
               show.legend = FALSE) +
  xlab(paste0("PC2 (", percentVar[2], "%)")) +
  ylab(paste0("PC3 (", percentVar[3], "%)")) +
  scale_color_manual(name = "social position", values = color_scheme) +
  scale_shape_manual(name = "clutch ID", values = shapes) +  # Specify shape values
  annotate("text", x = Inf, y = -Inf,                    #bottom right: Inf, y = -Inf, hjust = 1.1, vjust = -0.5
           label = "ADONIS p(social position) = 0.003",
           hjust = 1.1, 
           vjust = -0.5, 
           size = 4, fontface = "italic") +
  #ggtitle("PCA (PC2 vs PC3) all genes prcomp") +
  theme_classic() +
  theme(axis.text = element_text(size = 16), 
        axis.title = element_text(size = 16),         
        legend.title=element_text(size=13),  
        legend.text=element_text(size=13), legend.position = c(0.09,0.1))

pca_plot <- pca_plot + labs(tag = "A")+  # Add tag "A" to the plot
          theme(plot.tag = element_text(size = 16, face = "bold"))  # Adjust tag size and style

ggsave("../Figures/PCA_PC2_vs_PC3_all_genes_prcomp20260507.png", width = 10, height = 8, dpi = 1200)


## PERMANOVA on PCA 2 and PCA 3 
# Extract PC2 and PC3 for PERMANOVA
pca_matrix <- pca_df[, c("PC2", "PC3")]
# Run adonis PERMANOVA with fish_type2 as grouping variable
adonis_res <- adonis2(pca_matrix ~ fish_type2 + clutch_ID, data = pca_df, method = "euclidean", permutations = 999)
print(adonis_res)

#Permutation test for adonis under reduced model
#Terms added sequentially (first to last)
#Permutation: free
#Number of permutations: 999

#adonis2(formula = pca_matrix ~ fish_type2 + clutch_ID, data = pca_df, permutations = 999, method = "euclidean")
#           Df SumOfSqs      R2      F Pr(>F)   
#fish_type2  2   3531.9 0.21476 5.8911  0.003 **
#clutch_ID   2    923.0 0.05613 1.5395  0.187   
#Residual   40  11990.6 0.72911                 
#Total      44  16445.5 1.00000                 
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


#supplementary version of this figure:
pca_df$sample <- rownames(pca_df)

library(ggrepel)

pca_df$sample <- rownames(pca_df)

pca_plot_names <- ggplot(pca_df, aes(x = PC2, y = PC3, color = fish_type2)) +
            geom_point(size = 5, alpha = 0.8) +
            geom_text_repel(aes(label = sample),
            size = 3.5,
            box.padding = 0.35,
            point.padding = 0.3,
            segment.size = 0.2,
            max.overlaps = Inf) +    # set a large value or remove in older ggrepel
            stat_ellipse(geom = "polygon", aes(colour = fish_type2),
            fill = NA, type = "t", level = 0.95, alpha = 0.2, size= 2,
            show.legend = FALSE) +
            xlab(paste0("PC2 (", percentVar[2], "%)")) +
            ylab(paste0("PC3 (", percentVar[3], "%)")) +
            scale_color_manual(name = "social position", values = color_scheme) +
            annotate("text", x = Inf, y = -Inf,
            label = "ADONIS p(social position) = 0.003",
            hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic") +
            theme_classic() +
            theme(axis.text = element_text(size = 16),
            axis.title = element_text(size = 16),
            legend.title=element_text(size=13),
            legend.text=element_text(size=13),
            legend.position = c(0.09,0.1)) +
            labs(tag = "A") +
            theme(plot.tag = element_text(size = 16, face = "bold"))

ggsave("../Figures/PCA_PC2_vs_PC3_all_genes_prcomp_wnames_20260507.png",pca_plot_names, width = 10, height = 8, dpi = 1200)
ggsave("../Figures/PCA_PC2_vs_PC3_all_genes_prcomp_wnames_20260507.svg",pca_plot_names, width = 10, height = 8, dpi = 1200)

library(dplyr)
library(ggrepel)   # optional, for non-overlapping labels
library(ggplot2)

# Make sample column from rownames
pca_df1 <- pca_df %>% mutate(sample = rownames(pca_df))

# Extract numeric ID shared by P1/P2/S (adjust regex if needed)
# This assumes sample names start with digits like "10P1", "10P2", "10S", or contain the number at start.
pca_df1 <- pca_df1 %>% mutate(pair_id = sub("^([0-9]+).*", "\\1", sample))

# Ensure pca_df1 is a data.frame
pca_df1 <- as.data.frame(pca_df1)

# Optionally keep only IDs that have at least 2 or 3 samples (here >=2)
pair_counts <- pca_df1 %>% count(pair_id)
valid_pairs <- pair_counts %>% filter(n >= 2) %>% pull(pair_id)
pca_df_pairs <- pca_df1 

# For consistent line ordering, decide on an ordering variable within each pair.
# For a smooth visual, order by PC2 (or PC1). Create an "order" column per pair.
pca_df_pairs <- pca_df_pairs %>%
  group_by(pair_id) %>%
  arrange(PC2, .by_group = TRUE) %>%    # change to arrange(PC1, .by_group = TRUE) if preferred
  mutate(pt_order = row_number()) %>%
  ungroup()

# Plot: draw lines connecting all samples that share pair_id, then points on top
pca_plot_pairs <- ggplot() +
  # lines connecting the groups (use pca_df_pairs)
  geom_line(data = pca_df_pairs,
            aes(x = PC2, y = PC3, group = pair_id),
            color = "black", alpha = 0.6, size = 0.8) +
  # optional: show a point at each sample
  geom_point(data = pca_df1,
             aes(x = PC2, y = PC3, color = fish_type2, shape = clutch_ID),
             size = 4, alpha = 0.9) +
  # optional: add labels (uncomment if needed)
  geom_text_repel(data = pca_df, aes(x = PC2, y = PC3, label = sample, color = fish_type2),
                   size = 3.0, box.padding = 0.3, point.padding = 0.2, segment.size = 0.2) +
  stat_ellipse(data = pca_df1, aes(x = PC2, y = PC3, colour = fish_type2),
               fill = NA, type = "t", level = 0.95, alpha = 0.8, size = 1.5, show.legend = FALSE) +
  xlab(paste0("PC2 (", percentVar[2], "%)")) +
  ylab(paste0("PC3 (", percentVar[3], "%)")) +
  scale_color_manual(name = "social position", values = color_scheme) +
  scale_shape_manual(name = "clutch ID", values = shapes) +
  theme_classic() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 11),
        legend.position = c(0.09, 0.18))

# Add tag and ADONIS label if desired (example)
p_label <- "ADONIS p(social position) = 0.003"
pca_plot_pairs <- pca_plot_pairs +
  annotate("text", x = Inf, y = -Inf, label = p_label,
           hjust = 1.02, vjust = -0.2, size = 4, fontface = "italic") +
  #labs(tag = "A") +
  theme(plot.tag = element_text(size = 16, face = "bold"))

# Save
ggsave("../Figures/PCA_PC2_vs_PC3_with_pairs.svg",pca_plot_pairs, width = 10, height = 8, dpi = 1200)
ggsave("../Figures/PCA_PC2_vs_PC3_with_pairs.png",pca_plot_pairs, width = 10, height = 8, dpi = 1200)
```

```{r #PCA - PC1&PC3}

## Plot PCA (PC1 vs PC3) - manuscript version supplements
pca_plot13 <- ggplot(pca_df, aes(x = PC1, y = PC3, color = fish_type2)) +
  geom_point(size = 5, aes(shape = clutch_ID), alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2), 
               fill = NA, type = "t", level = 0.95, alpha = 0.2, size = 2,
               show.legend = FALSE) +
  xlab(paste0("PC1 (", percentVar[1], "%)")) +
  ylab(paste0("PC3 (", percentVar[3], "%)")) +
  scale_color_manual(name = "social position", values = color_scheme) +
  scale_shape_manual(name = "clutch ID", values = shapes) +
  annotate("text", x = Inf, y = -Inf,                    #bottom right: Inf, y = -Inf, hjust = 1.1, vjust = -0.5
            label = "ADONIS p(social position) = 0.5",
           hjust = 1.1, vjust = -0.5, 
           size = 4.5, fontface = "italic") +
  theme_classic() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 11),
        legend.position = c(0.07, 0.18))

ggsave("../Figures/PCA_PC1_vs_PC3_all_genes_prcomp.png", width = 10, height = 8, dpi = 1200)

# PERMANOVA on PCA 1 and PCA 3
pca_matrix_pc1_pc3 <- pca_df[, c("PC1", "PC3")]
adonis_res_pc1_pc3 <- adonis2(pca_matrix_pc1_pc3 ~ fish_type2 + clutch_ID, 
                               data = pca_df, 
                               method = "euclidean", 
                               permutations = 999,
                               by = "margin")   # <-- add this
print(adonis_res_pc1_pc3)

#Permutation test for adonis under reduced model
#Marginal effects of terms
#Permutation: free
#Number of permutations: 999

#adonis2(formula = pca_matrix_pc1_pc3 ~ fish_type2 + clutch_ID, data = pca_df, permutations = 999, method = "euclidean", by = "margin")
#           Df SumOfSqs      R2      F Pr(>F)  
#fish_type2  2    762.4 0.03165 0.7348  0.553  
#clutch_ID   2   2388.7 0.09915 2.3021  0.071 .
#Residual   40  20752.6 0.86142                
#Total      44  24091.3 1.00000                
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


```


```{r supplements PCA figure combined}
#Combine PCA plots into one figure for supplements (PC1 vs PC2 and PC1 vs PC3)
library(cowplot)
combined_pca_plot <- plot_grid(pca_plot12, pca_plot13, 
                                labels = c("A", "B"),
                                label_size = 16,
                                ncol = 2)
ggsave("../Figures/combined_PCA_12and13_plots.png", combined_pca_plot, width = 19, height = 8, dpi = 1200)
```


```{r}
# ============================================================
# PCA — all genes, top 50%, 25%, and 5% most variable genes
# ============================================================
library(ggplot2)
library(patchwork)
library(vegan)      # for adonis2

gene_var    <- apply(pca_data_all, 1, var)
total_genes <- nrow(pca_data_all)

color_scheme <- c("magenta3", "purple3", "grey50")
shapes <- c(15,  16,  17)

cat("Total genes:", total_genes, "\n")
cat("Top 50%:",  round(total_genes * 0.50), "genes\n")
cat("Top 25%:",  round(total_genes * 0.25), "genes\n")
cat("Top  5%:",  round(total_genes * 0.05), "genes\n")

# ============================================================
# Function: run PCA
# ============================================================
run_pca <- function(expr_mat, top_percent, vsd_obj) {
  n_genes    <- round(nrow(expr_mat) * top_percent)
  top_genes  <- names(sort(gene_var, decreasing = TRUE))[1:n_genes]
  subset_mat <- expr_mat[top_genes, ]

  pca_res <- prcomp(t(subset_mat), center = TRUE, scale. = FALSE)

  pca_df <- as.data.frame(pca_res$x[, 1:3])
  pca_df$fish_type2 <- colData(vsd_obj)$fish_type2
  pca_df$clutch_ID  <- colData(vsd_obj)$clutch_ID

  percentVar <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

  list(pca_df = pca_df, percentVar = percentVar, n_genes = n_genes)
}

# ============================================================
# Function: run PERMANOVA on two PC axes, return fish_type2 p-value
# ============================================================
run_adonis <- function(pca_obj, pcx, pcy) {
  df         <- pca_obj$pca_df
  pca_matrix <- df[, c(pcx, pcy)]

  adonis_res <- adonis2(pca_matrix ~ fish_type2 + clutch_ID,
                        data         = df,
                        method       = "euclidean",
                        permutations = 999,
                        by           = "margin")

  p_val <- adonis_res["fish_type2", "Pr(>F)"]
  return(p_val)
}

# ============================================================
# Helper: format p-value for annotation
# ============================================================
format_pval <- function(p) {
  if (is.na(p)) return("ADONIS p(social position) = NA")
  if (p < 0.001) return("ADONIS p(social position) < 0.05")
  paste0("ADONIS p(social position) = ", round(p, 3))
}

# ============================================================
# Run PCA for every gene subset
# ============================================================
pca_all <- run_pca(pca_data_all, 1.00, vsd)
pca_50  <- run_pca(pca_data_all, 0.50, vsd)
pca_25  <- run_pca(pca_data_all, 0.25, vsd)
pca_05  <- run_pca(pca_data_all, 0.05, vsd)

# ============================================================
# Run PERMANOVA for all 12 combinations
# ============================================================
cat("Running PERMANOVA for all 12 PC combinations...\n")

# All genes
pval_all_12 <- run_adonis(pca_all, "PC1", "PC2")
pval_all_23 <- run_adonis(pca_all, "PC2", "PC3")
pval_all_13 <- run_adonis(pca_all, "PC1", "PC3")

# Top 50%
pval_50_12  <- run_adonis(pca_50,  "PC1", "PC2")
pval_50_23  <- run_adonis(pca_50,  "PC2", "PC3")
pval_50_13  <- run_adonis(pca_50,  "PC1", "PC3")

# Top 25%
pval_25_12  <- run_adonis(pca_25,  "PC1", "PC2")
pval_25_23  <- run_adonis(pca_25,  "PC2", "PC3")
pval_25_13  <- run_adonis(pca_25,  "PC1", "PC3")

# Top 5%
pval_05_12  <- run_adonis(pca_05,  "PC1", "PC2")
pval_05_23  <- run_adonis(pca_05,  "PC2", "PC3")
pval_05_13  <- run_adonis(pca_05,  "PC1", "PC3")

cat("PERMANOVA complete.\n")

# ============================================================
# Function: make PCA plot with PERMANOVA p-value annotation
# ============================================================
make_pca_plot <- function(pca_obj, pcx, pcy, p_val) {

  df     <- pca_obj$pca_df
  pVar   <- pca_obj$percentVar
  pc_idx <- as.integer(gsub("PC", "", c(pcx, pcy)))
  label  <- format_pval(p_val)

  ggplot(df, aes_string(x = pcx, y = pcy, color = "fish_type2")) +
    geom_point(size = 5, aes(shape = clutch_ID), alpha = 0.8) +
    stat_ellipse(geom = "polygon", aes(colour = fish_type2),
                 fill = NA, type = "t", level = 0.95,
                 alpha = 0.2, size = 2, show.legend = FALSE) +
    xlab(paste0(pcx, " (", pVar[pc_idx[1]], "%)")) +
    ylab(paste0(pcy, " (", pVar[pc_idx[2]], "%)")) +
    scale_color_manual(name = "social position", values = color_scheme) +
    scale_shape_manual(name = "clutch ID",        values = shapes) +
    annotate("text",
             x = Inf, y = -Inf,
             label = label,
             hjust = 1.05, vjust = -0.5,
             size = 3.5, fontface = "italic") +
    theme_classic() +
    theme(axis.text    = element_text(size = 14),
          axis.title   = element_text(size = 14),
          legend.title = element_text(size = 12),
          legend.text  = element_text(size = 12),
          plot.tag     = element_text(size = 14, face = "bold"))
}

# ============================================================
# Generate all 12 plots with unique tags A–L
# ============================================================

# Row 1 — all genes
pAll_12 <- make_pca_plot(pca_all, "PC1", "PC2", pval_all_12) + labs(tag = "A")
pAll_23 <- make_pca_plot(pca_all, "PC2", "PC3", pval_all_23) + labs(tag = "B")
pAll_13 <- make_pca_plot(pca_all, "PC1", "PC3", pval_all_13) + labs(tag = "C")

# Row 2 — top 50%
p50_12  <- make_pca_plot(pca_50,  "PC1", "PC2", pval_50_12)  + labs(tag = "D")
p50_23  <- make_pca_plot(pca_50,  "PC2", "PC3", pval_50_23)  + labs(tag = "E")
p50_13  <- make_pca_plot(pca_50,  "PC1", "PC3", pval_50_13)  + labs(tag = "F")

# Row 3 — top 25%
p25_12  <- make_pca_plot(pca_25,  "PC1", "PC2", pval_25_12)  + labs(tag = "G")
p25_23  <- make_pca_plot(pca_25,  "PC2", "PC3", pval_25_23)  + labs(tag = "H")
p25_13  <- make_pca_plot(pca_25,  "PC1", "PC3", pval_25_13)  + labs(tag = "I")

# Row 4 — top 5%
p05_12  <- make_pca_plot(pca_05,  "PC1", "PC2", pval_05_12)  + labs(tag = "J")
p05_23  <- make_pca_plot(pca_05,  "PC2", "PC3", pval_05_23)  + labs(tag = "K")
p05_13  <- make_pca_plot(pca_05,  "PC1", "PC3", pval_05_13)  + labs(tag = "L")

# ============================================================
# Row and column label helpers
# ============================================================
make_row_label <- function(label) {
  ggplot() +
    annotate("text", x = 0.5, y = 0.5,
             label = label, angle = 90,
             size = 5, fontface = "bold") +
    theme_void()
}

make_col_label <- function(label) {
  ggplot() +
    annotate("text", x = 0.5, y = 0.5,
             label = label,
             size = 5, fontface = "bold") +
    theme_void()
}

corner <- make_col_label("")
c1_h   <- make_col_label("PC1 vs PC2")
c2_h   <- make_col_label("PC2 vs PC3")
c3_h   <- make_col_label("PC1 vs PC3")

r1_l   <- make_row_label("All genes")
r2_l   <- make_row_label("Top 50%")
r3_l   <- make_row_label("Top 25%")
r4_l   <- make_row_label("Top 5%")

# ============================================================
# Assemble 5 × 4 grid
# ============================================================
design <- "
ABCD
EFGH
IJKL
MNOP
QRST
"

combined_all <- wrap_plots(
  corner,  c1_h,    c2_h,    c3_h,
  r1_l,    pAll_12, pAll_23, pAll_13,
  r2_l,    p50_12,  p50_23,  p50_13,
  r3_l,    p25_12,  p25_23,  p25_13,
  r4_l,    p05_12,  p05_23,  p05_13,
  design  = design,
  widths  = c(0.06, 1, 1, 1),
  heights = c(0.06, 1, 1, 1, 1),
  guides  = "collect"
)

# ============================================================
# Save
# ============================================================
#ggsave("../Figures/PCA_all_50_25_5_combined.pdf",
#       combined_all,
#       width  = 22,
#       height = 28,
#       device = cairo_pdf)

ggsave("../Figures/PCA_all_50_25_5_combined_pval.png",
       combined_all,
       width  = 22,
       height = 28,
       dpi    = 300)
```



```{r}
### PCA plot using DESeq2 package of ntop 500 genes - very stringent way (so not ideal explores top ~2.5% of genes)
### ntop = 500 is the top 500 genes with the highest variance across samples

PCAdata_500 <- plotPCA(vsd, intgroup=c("clutch_ID", "fish_type2"), ntop=500, returnData=TRUE)
percentVar_500 <- round(100 * attr(PCAdata_500, "percentVar"))
PCAdata_500

pca_plot_500 <- ggplot(PCAdata_500, aes(x = PC1, y = PC2, color = fish_type2)) +
  geom_point(size = 3, aes(fill = fish_type2, shape=clutch_ID), alpha = 0.8) +
  stat_ellipse(geom = "polygon", aes(colour = fish_type2, fill = fish_type2), fill = NA, type = "t", level = 0.95, alpha = 0.2, show.legend = FALSE) +
  #geom_text(aes(label = name), vjust = -0.5, hjust = 0.5, size = 2, color = "black") +  # Add this line to include sample names as labels+
  xlab(paste0("PC1 (", percentVar_500[1], "%)")) +
  scale_fill_manual(name = "Treatment", values=color_scheme) +
  scale_color_manual(name = "Treatment", values=color_scheme) +
  ylab(paste0("PC2 (", percentVar_500[2], "%)")) +
  ggtitle("Top 500 genes") +
  theme_classic()
pca_plot_500

```


## Differential gene expression {.tabset}

<br/>

### DEGs 

```{r general results of DEGs}
#Look at results extracts as a table from a DESeq analysis giving base means across samples, log2 fold changes, standard errors, test statistics, p-values and adjusted p-values; resultsNames returns the names of the estimated effects (coefficents) of the model
resultsNames(dds)

## Fist, we will look at the general 'results()' function
results(dds)
summary(results(dds))
```



```{r create results of DEGs by contrasts}
load("transformed_counts_mindepth10_run2.RData")
## When specifying your contrasts, make sure you 'control' level is the second value input and your 'treatment' is the first
P1_con_P2 <- results(dds, contrast = c("fish_type2", "P2", "P1"), alpha = 0.05)
P1_con_S <- results(dds, contrast = c("fish_type2", "S", "P1"), alpha = 0.05)
P2_con_S <- results(dds, contrast = c("fish_type2", "S", "P2"), alpha = 0.05)


#Contrasting clutches
L2_con_L3 <- results(dds, contrast = c("clutch_ID", "L2", "L3"), alpha = 0.05)
L2_con_L6 <- results(dds, contrast = c("clutch_ID", "L2", "L6"), alpha = 0.05)
L3_con_L6 <- results(dds, contrast = c("clutch_ID", "L3", "L6"), alpha = 0.05)

```

```{r results summary, echo = TRUE}

summary(P1_con_P2) #  P1 individuals compared to P2
summary(P1_con_S) # P1 individuals compared to S
summary(P2_con_S) # P2 individuals comapred to S


summary(L2_con_L3)  #L2 indiviuals compared to L3 clutch
summary(L2_con_L6)  #L2 indiviuals compared to L6 clutch
summary(L3_con_L6)  #L3 indiviuals compared to L6 clutch

```


```{r filter for DEGs by treatment}

## Filter for DEG per pairwise comparison (adjusted p value = 0.05)
# P1 v P2
downCon_P1P2 <- row.names(P1_con_P2[P1_con_P2$padj < 0.05 & !is.na(P1_con_P2$padj) & P1_con_P2$log2FoldChange < 0, ])
upCon_P1P2 <- row.names(P1_con_P2[P1_con_P2$padj < 0.05 & !is.na(P1_con_P2$padj) & P1_con_P2$log2FoldChange > 0, ])
allCon_P1P2 <- row.names(P1_con_P2[P1_con_P2$padj < 0.05 & !is.na(P1_con_P2$padj), ])

# P1 vs S
downCon_P1S <- row.names(P1_con_S[P1_con_S$padj < 0.05 & !is.na(P1_con_S$padj) & P1_con_S$log2FoldChange < 0, ])
upCon_P1S <- row.names(P1_con_S[P1_con_S$padj < 0.05 & !is.na(P1_con_S$padj) & P1_con_S$log2FoldChange > 0, ])
allCon_P1S <- row.names(P1_con_S[P1_con_S$padj < 0.05 & !is.na(P1_con_S$padj), ])

# P2 vs S
downCon_P2S <- row.names(P2_con_S[P2_con_S$padj < 0.05 & !is.na(P2_con_S$padj) & P2_con_S$log2FoldChange < 0, ])
upCon_P2S <- row.names(P2_con_S[P2_con_S$padj < 0.05 & !is.na(P2_con_S$padj) & P2_con_S$log2FoldChange > 0, ])
allCon_P2S <- row.names(P2_con_S[P2_con_S$padj < 0.05 & !is.na(P2_con_S$padj), ]) 


## Compile DEG dataframe
DEG_bar <- data.frame("fish_type" = c(toString(unlist(str_split(P1_con_P2@elementMetadata@listData[["description"]][2], " "))[c(8,6)]), toString(unlist(str_split(P1_con_S@elementMetadata@listData[["description"]][2], " "))[c(8,6)]), toString(unlist(str_split(P2_con_S@elementMetadata@listData[["description"]][2], " "))[c(8,6)])), 
                "up" = c(length(upCon_P1P2), length(upCon_P1S), length(upCon_P2S)), 
                "down" = c((length(downCon_P1P2) * -1), (length(downCon_P1S) * -1), (length(downCon_P2S) * -1)))


DEG_bar <- gather(DEG_bar, reg, genes, up:down)
DEG_bar
```


```{r filter for DEGs by clutch}
#Doing the same for clutches

## Filter for DEG per pairwise comparison (adjusted p value = 0.05)
# L2 vs L3
downCon_L2L3 <- row.names(L2_con_L3[L2_con_L3$padj < 0.05 & !is.na(L2_con_L3$padj) & L2_con_L3$log2FoldChange < 0, ])
upCon_L2L3 <- row.names(L2_con_L3[L2_con_L3$padj < 0.05 & !is.na(L2_con_L3$padj) & L2_con_L3$log2FoldChange > 0, ])
allCon_L2L3 <- row.names(L2_con_L3[L2_con_L3$padj < 0.05 & !is.na(L2_con_L3$padj), ])

# L2 vs L6
downCon_L2L6 <- row.names(L2_con_L6[L2_con_L6$padj < 0.05 & !is.na(L2_con_L6$padj) & L2_con_L6$log2FoldChange < 0, ])
upCon_L2L6 <- row.names(L2_con_L6[L2_con_L6$padj < 0.05 & !is.na(L2_con_L6$padj) & L2_con_L6$log2FoldChange > 0, ])
allCon_L2L6 <- row.names(L2_con_L6[L2_con_L6$padj < 0.05 & !is.na(L2_con_L6$padj), ])

# L3 vs L6
downCon_L3L6 <- row.names(L3_con_L6[L3_con_L6$padj < 0.05 & !is.na(L3_con_L6$padj) & L3_con_L6$log2FoldChange < 0, ])
upCon_L3L6 <- row.names(L2_con_L6[L2_con_L6$padj < 0.05 & !is.na(L2_con_L6$padj) & L2_con_L6$log2FoldChange > 0, ])
allCon_L3L6 <- row.names(L3_con_L6[L3_con_L6$padj < 0.05 & !is.na(L3_con_L6$padj), ]) 


## Compile DEG dataframe
DEG_bar_clutches <- data.frame("clutch_ID" = c(toString(unlist(str_split(L2_con_L3@elementMetadata@listData[["description"]][2], " "))[c(8,6)]), toString(unlist(str_split(L2_con_L6@elementMetadata@listData[["description"]][2], " "))[c(8,6)]), toString(unlist(str_split(L3_con_L6@elementMetadata@listData[["description"]][2], " "))[c(8,6)])), 
                "up" = c(length(upCon_L2L3), length(upCon_L2L6), length(upCon_L3L6)), 
                "down" = c((length(downCon_L2L3) * -1), (length(downCon_L2L6) * -1), (length(downCon_L3L6) * -1)))


DEG_bar_clutches <- gather(DEG_bar_clutches, reg, genes, up:down)
#DEG_bar_clutches
```





```{r DEG plot}

## Plot bar graph of DEG by treatment

ggplot(data = DEG_bar, aes(x = fish_type, y = genes, fill = reg)) +
  theme_classic() +
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("#2c7bb6", "#d7191c")) +
  geom_hline(yintercept = 0, linetype = "dashed", size=0.5) + 
  ylab("Sig DEGs by treatment") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 90)) +
  xlab ("")

## Plot bar graph of DEG by clutch

ggplot(data = DEG_bar_clutches, aes(x = clutch_ID, y = genes, fill = reg)) +
  theme_classic() +
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("#2c7bb6", "#d7191c")) +
  geom_hline(yintercept = 0, linetype = "dashed", size=0.5) + 
  ylab("Sig DEGs by clutch") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 90)) +
  xlab ("")

```


Figure: Bar plot of significant DEGs. The y axis denotes how many genes there are, with positive values being up-regulated genes, and negative values denoting down-regulated genes. 


<br/>

### Paired Venns of DEGs


```{r paired Venns}

## Comparing up between all three: up regulation
venn_loci_up <- list("P1 vs P2" = upCon_P1P2, "P1 vs S" = upCon_P1S, "P2 vs S" = upCon_P2S)

## Comparing up between all three: down regulation
venn_loci_down <- list("P1 vs P2" = downCon_P1P2, "P1 vs S" = downCon_P1S, "P2 vs S" = downCon_P2S)

## Comparing between all three
venn_all <- list("P1 vs P2" = allCon_P1P2, "P1 vs S" = allCon_P1S, "P2 vs S" = allCon_P2S)

ven_all_up <- ggvenn(venn_loci_up, 
       show_percentage=FALSE,
       fill_color = c("darkorange", "coral", "#C8322F"), fill_alpha = 0.3,
       stroke_size = 0.4, text_size = 3, text_color = "#25292B",
       set_name_color = "#25292B", set_name_size = 2
       )
ven_all_up
#ggsave( "../Figures/All_PairWise_Vens_UP.png", width = 4, height = 4)


ven_all_down <- ggvenn(venn_loci_down, 
       show_percentage=FALSE,
       fill_color = c("cyan4", "cornflowerblue", "blue"), fill_alpha = 0.3,
       stroke_size = 0.4, text_size = 3, text_color = "#25292B",
       set_name_color = "#25292B", set_name_size = 2
       )
ven_all_down       
#ggsave("../Figures/All_PairWise_Vens_DOWN.png", width = 4, height = 4)      



#read in geneID2geneName tab to match gene names with
gene_map <- read.delim("./geneID2geneNAME.tab", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

#down regulated genes
shared_genes_down_partial <- intersect(
  venn_loci_down[["P1 vs P2"]],
  venn_loci_down[["P1 vs S"]])

shared_genes_down_partial <- data.frame(
  ensembl_gene_id = shared_genes_down_partial,
  external_gene_name = gene_map$external_gene_name[match(shared_genes_down_partial, gene_map$ensembl_gene_id)])
# Save to CSV
#write.csv(shared_genes_down_partial, "../Data/VENN_shared_genes_down.csv", row.names = FALSE, quote = FALSE)

#up regulated genes
shared_genes_up_partial <- intersect(
  venn_loci_up[["P1 vs P2"]],
  venn_loci_up[["P1 vs S"]])

shared_genes_up_partial <- data.frame(
  ensembl_gene_id = shared_genes_up_partial,
  external_gene_name = gene_map$external_gene_name[match(shared_genes_up_partial, gene_map$ensembl_gene_id)])
# Save to CSV
#write.csv(shared_genes_up_partial, "../Data/VENN_shared_genes_up.csv", row.names = FALSE, quote = FALSE)

```

Figure 4: Venn diagrams depicting number of shared significantly up-regulated / down-regulated genes. Orange venn diagrams depict up-regulated genes, while blue depicts down-regulated genes. 


## Heatmaps {.tabset}


```{r heatmap of all annotated genes P1 vs P2 vs S}
## Load previously saved dds, vsd, and rlog data 
load("transformed_counts_mindepth10.RData")

#the three pairwise contrasts:
P1_con_P2 <- results(dds, contrast = c("fish_type2", "P2", "P1"), alpha = 0.05)
nrow(P1_con_P2[P1_con_P2$padj<0.05 & !is.na(P1_con_P2$padj),])  # Num significantly differentially expressed genes excluding the no/low count genes   #133
plotMA(P1_con_P2, main="P1 vs P2")

P1_con_S <- results(dds, contrast = c("fish_type2", "S", "P1"), alpha = 0.05)
nrow(P1_con_S[P1_con_S$padj<0.05 & !is.na(P1_con_S$padj),])  # Num significantly differentially expressed genes excluding the no/low count genes   #520
plotMA(P1_con_S, main="P1 vs S")


P2_con_S <- results(dds, contrast = c("fish_type2", "S", "P2"), alpha = 0.05)
nrow(P2_con_S[P2_con_S$padj<0.05 & !is.na(P2_con_S$padj),])  # Num significantly differentially expressed genes excluding the no/low count genes   #0
plotMA(P2_con_S, main="P2 vs S")


######-------getting p-vals & p-adjusted from pairwise contrasts

#P1 vs P2   #133
valP1vsP2=cbind(P1_con_P2$pvalue, P1_con_P2$padj)
head(valP1vsP2)
nrow(valP1vsP2)
colnames(valP1vsP2)=c("pval.P1vsP2", "padj.P1vsP2")
length(valP1vsP2[,1])
table(complete.cases(valP1vsP2))


#P1 vs S
valP1vsS=cbind(P1_con_S$pvalue, P1_con_S$padj)
head(valP1vsS)
colnames(valP1vsS)=c("pval.P1vsS", "padj.P1vsS")
length(valP1vsS[,1])
table(complete.cases(valP1vsS))

#P2 vs S
valP2vsS=cbind(P2_con_S$pvalue, P2_con_S$padj)
head(valP2vsS)
colnames(valP2vsS)=c("pval.P2vsS", "padj.P2vsS")
length(valP2vsS[,1])
table(complete.cases(valP2vsS))

######-------------make rlogdata and pvals table

rlog=rlogTransformation(dds, blind=TRUE) 
rld=assay(rlog)
head(rld)
length(rld[,1])

rldpvals=cbind(rld,valP1vsP2, valP1vsS, valP2vsS)
head(rldpvals)
dim(rldpvals)
# [1] 20537    51
table(complete.cases(rldpvals))
#FALSE  TRUE 
# 6372 14165 

#write.csv(rldpvals, "../Data/percula_RLDandPVALS.csv", quote=F)
```


basic heatmap with up and downregulated genes across 3 fish categories (social positions)
```{r}
rldpvals <- read.csv(file="../Data/percula_RLDandPVALS.csv", row.names=1)
head(rldpvals)


p.val=0.05 # FDR cutoff
conds=rldpvals[rldpvals$padj.P1vsP2<=p.val & !is.na(rldpvals$padj.P1vsP2) & rldpvals$padj.P1vsS<=p.val & !is.na(rldpvals$padj.P1vsS),]
length(conds[,1])
#308 pval=0.1 when only P1vsP2
#251 pval=0.1 when both P1vsP2 & P1vsS
#109 pval=0.05when both P1vsP2 & P1vsS
head(conds)

# Remove statistical columns at the end (those starting with "pval" or "padj")
expr_clean <- conds[, !grepl("^pval|^padj", colnames(conds))]

#remove X from column names
colnames(expr_clean) <- gsub("^X", "", colnames(expr_clean))

fish_type <- c(rep("S", 15), rep("P1", 15), rep("P2", 15))
# Reorder columns by fish type
ordered_indices <- order(fish_type)
expr_ordered <- expr_clean[, ordered_indices]
fish_type_ordered <- fish_type[ordered_indices]

# Create annotation dataframe
annotation_col <- data.frame(FishType = fish_type_ordered)
rownames(annotation_col) <- colnames(expr_ordered)

# Define annotation colors
ann_colors <- list(
  FishType = c(S = "grey", P1 = "magenta2", P2 = "purple3")
)

col0 <- colorRampPalette(rev(c("chocolate1", "#FEE090", "grey10", "cyan3", "cyan")))(100)

# Plot the heatmap
pheatmap(as.matrix(expr_clean),  # Exclude the first four columns (gene names and p-values)
         scale = "row",
         annotation_col = annotation_col,
         annotation_colors = ann_colors,
         cluster_cols = FALSE,         # disables clustering to preserve group order
         show_colnames = T,
         show_rownames = FALSE,        # hides gene names
         fontsize_row = 8,
         color = col0)

#ggsave("../Figures/Heatmap_P1_vs_P2_vs_S.png", width = 10, height = 8, dpi = 1200)
```

This heatmap shows the expression of genes that are significantly differentially expressed between the three social positions (P1, P2, and S). The rows represent genes, while the columns represent individuals that belong to the 3 different social positions. The color intensity indicates the level of gene expression, with red indicating higher expression and blue indicating lower expression.
There seem to be subset of DEGs. To further explore these DEGs and link DEGs with observed phenotypes, I will explore this dataset using WGCNA_script.R, which will allow me to identify modules of co-expressed genes and link these modules to phenotypes.

#### Vulcano Plots for Supplementary Materials

```{r vulcano plot - of pairwise comparisons}
## Load previously saved dds, vsd, and rlog data 
load("transformed_counts_mindepth10.RData")

#P1 vs S
res_P1vsS <- results(dds, contrast = c("fish_type2", "P1", "S"))
res_df_P1vsS <- as.data.frame(res_P1vsS)
res_df_P1vsS$gene <- rownames(res_df_P1vsS)  # add gene names

# Define thresholds
log2fc_thresh <- 1
padj_thresh <- 0.05 

# recode categories - for significance
res_df_P1vsS$significance <- "NotSig"
res_df_P1vsS$significance[res_df_P1vsS$padj < padj_thresh &
                  res_df_P1vsS$log2FoldChange >=  log2fc_thresh] <- "UpinP1"
res_df_P1vsS$significance[res_df_P1vsS$padj < padj_thresh &
                  res_df_P1vsS$log2FoldChange <= -log2fc_thresh] <- "UpinS"

# keep a clean ordering (optional)
res_df_P1vsS$significance <- factor(
  res_df_P1vsS$significance,
  levels = c("UpinP1", "UpinS", "NotSig"))

custom_colors <- c(
  "UpinP1" = "magenta3",       
  "UpinS" = "gray40",
  "NotSig" = "black")

res_df_P1vsS$gene <- rownames(res_df_P1vsS)

# Top 10 upregulated in P1 (positive FC)
top_up_P1 <- res_df_P1vsS %>%
  filter(log2FoldChange >= log2fc_thresh & padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

# Top 10 upregulated in S (negative FC)
top_up_S <- res_df_P1vsS %>%
  filter(log2FoldChange <= -log2fc_thresh & padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

# Combine them into one dataframe
highlight_genes <- bind_rows(top_up_P1, top_up_S)
write.csv(highlight_genes, "log2FC_top10genes_P1vsS_2025_06_26.csv", row.names = FALSE)

P1vsS <- ggplot(res_df_P1vsS, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), size = 3.5, alpha = 0.7) +
  scale_color_manual(values = custom_colors) +
  geom_vline(xintercept = c(-log2fc_thresh, log2fc_thresh), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey80") +
  #geom_text_repel(aes(label = gene), size = 3.5, min.segment.length = 0) +
  labs(x = "Log2FC",
      y = "-log10 adj.p-value",
       color = "") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",  # remove legend if not needed
    panel.grid = element_blank(),
     axis.text = element_text(size = 14),      # increase tick label size
    axis.title = element_text(size = 16),
    axis.line = element_line(color = "black", linewidth = 0.5),
    plot.background = element_rect(fill = "white", color = NA))     



library(plotly)
library(htmlwidgets)

# Add tooltip info column
res_df_P1vsS$tooltip <- paste0(
  "Gene: ", res_df_P1vsS$gene, "\n",
  "log2FC: ", round(res_df_P1vsS$log2FoldChange, 2), "\n",
  "-log10(padj): ", round(-log10(res_df_P1vsS$padj), 2)
)

highlight_genes$tooltip <- paste0(
  "Gene: ", highlight_genes$gene, "\n",
  "log2FC: ", round(highlight_genes$log2FoldChange, 2), "\n",
  "-log10(padj): ", round(-log10(highlight_genes$padj), 2)
)

# Create ggplot with 'text' aesthetic
p <- ggplot(res_df_P1vsS, aes(x = log2FoldChange, y = -log10(padj), text = tooltip)) +
  geom_point(aes(color = significance), size = 2, alpha = 0.7) +
  geom_point(data = highlight_genes,
             aes(x = log2FoldChange, y = -log10(padj), text = tooltip),
             shape = 21, stroke = 1, color = "red", fill = NA, size = 3) +
  geom_vline(xintercept = c(-log2fc_thresh, log2fc_thresh), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey80") +
  labs(x = "Log2FC (P1 vs S)", y = "-log10 adj. p-value", color = "") +
  scale_color_manual(values = custom_colors) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

# Convert to interactive plot
P1vsS <- ggplotly(p, tooltip = "text")
saveWidget(P1vsS, file = "volcano_plot_P1_vs_S.html")
```
Positive log2FoldChange → gene is upregulated in P1 relative to S

Negative log2FoldChange → gene is downregulated in P1 relative to S
→ i.e., upregulated in S

```{r}
#P1 vs P2
res_P1vsP2 <- results(dds, contrast = c("fish_type2", "P1", "P2"))
res_df_P1vsP2 <- as.data.frame(res_P1vsP2)
res_df_P1vsP2$gene <- rownames(res_df_P1vsP2)  # add gene names

# Define thresholds
log2fc_thresh <- 1
padj_thresh <- 0.05 

# recode categories - for significance
res_df_P1vsP2$significance <- "NotSig"
res_df_P1vsP2$significance[res_df_P1vsP2$padj < padj_thresh &
                  res_df_P1vsP2$log2FoldChange >=  log2fc_thresh] <- "UpinP1"
res_df_P1vsP2$significance[res_df_P1vsP2$padj < padj_thresh &
                  res_df_P1vsP2$log2FoldChange <= -log2fc_thresh] <- "UpinP2"

# keep a clean ordering (optional)
res_df_P1vsP2$significance <- factor(
  res_df_P1vsP2$significance,
  levels = c("UpinP1", "UpinP2", "NotSig"))

custom_colors2 <- c(
  "UpinP1" = "magenta3",       
  "UpinP2" = "purple3",
  "NotSig" = "black")

res_df_P1vsP2$gene <- rownames(res_df_P1vsP2)

# Top 10 upregulated in P1 (positive FC)
top_up_P1 <- res_df_P1vsP2 %>%
  filter(log2FoldChange >= log2fc_thresh & padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

# Top 10 upregulated in P2 (negative FC)
top_up_P2 <- res_df_P1vsP2 %>%
  filter(log2FoldChange <= -log2fc_thresh & padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

# Combine them into one dataframe
highlight_genes <- bind_rows(top_up_P1, top_up_P2)
write.csv(highlight_genes, "log2FC_top10_genes_P1vsP2_2025_06_26.csv", row.names = FALSE)

P1vsP2 <- ggplot(res_df_P1vsP2, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), size = 3.5, alpha = 0.7) +
  scale_color_manual(values = custom_colors2) +
  geom_vline(xintercept = c(-log2fc_thresh, log2fc_thresh), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey80") +
  #geom_text_repel(aes(label = gene), size = 3.5, min.segment.length = 0) +
  labs(x = "Log2FC",
      y = "-log10 adj.p-value",
       color = "") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",  # remove legend if not needed
    panel.grid = element_blank(),
     axis.text = element_text(size = 14),      # increase tick label size
    axis.title = element_text(size = 16),
    axis.line = element_line(color = "black", linewidth = 0.5),
    plot.background = element_rect(fill = "white", color = NA))     


# Add tooltip info column
res_df_P1vsP2$tooltip <- paste0(
  "Gene: ", res_df_P1vsP2$gene, "\n",
  "log2FC: ", round(res_df_P1vsP2$log2FoldChange, 2), "\n",
  "-log10(padj): ", round(-log10(res_df_P1vsP2$padj), 2)
)

highlight_genes$tooltip <- paste0(
  "Gene: ", highlight_genes$gene, "\n",
  "log2FC: ", round(highlight_genes$log2FoldChange, 2), "\n",
  "-log10(padj): ", round(-log10(highlight_genes$padj), 2)
)

# Create ggplot with 'text' aesthetic
p <- ggplot(res_df_P1vsP2, aes(x = log2FoldChange, y = -log10(padj), text = tooltip)) +
  geom_point(aes(color = significance), size = 2, alpha = 0.7) +
  geom_point(data = highlight_genes,
             aes(x = log2FoldChange, y = -log10(padj), text = tooltip),
             shape = 21, stroke = 1, color = "red", fill = NA, size = 3) +
  geom_vline(xintercept = c(-log2fc_thresh, log2fc_thresh), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey80") +
  labs(x = "Log2FC (P1 vs P2)", y = "-log10 adj. p-value", color = "") +
  scale_color_manual(values = custom_colors) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

# Convert to interactive plot
P1vsP2 <- ggplotly(p, tooltip = "text")
saveWidget(P1vsP2, file = "volcano_plot_P1_vs_P2.html")
```


```{r}
#P2 vs S
res_P2vsS <- results(dds, contrast = c("fish_type2", "P2", "S"))
res_df_P2vsS <- as.data.frame(res_P2vsS)
res_df_P2vsS$gene <- rownames(res_df_P2vsS)  # add gene names

# Define thresholds
log2fc_thresh <- 1
padj_thresh <- 0.05

# recode categories - for significance
res_df_P2vsS$significance <- "NotSig"
res_df_P2vsS$significance[res_df_P2vsS$padj < padj_thresh &
                  res_df_P2vsS$log2FoldChange >=  log2fc_thresh] <- "UpinP2"
res_df_P2vsS$significance[res_df_P2vsS$padj < padj_thresh &
                  res_df_P2vsS$log2FoldChange <= -log2fc_thresh] <- "UpinS"

# keep a clean ordering (optional)
res_df_P2vsS$significance <- factor(
  res_df_P2vsS$significance,
  levels = c("UpinP2", "UpinS", "NotSig"))

custom_colors3 <- c(
  "UpinP2" = "purple3",       
  "UpinS" = "gray40",
  "NotSig" = "black")

res_df_P2vsS$gene <- rownames(res_df_P2vsS)


# Top 10 upregulated in P2 (positive FC)
top_up_P2 <- res_df_P2vsS %>%
  filter(log2FoldChange >= log2fc_thresh & padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

# Top 10 upregulated in S (negative FC)
top_up_S <- res_df_P2vsS %>%
  filter(log2FoldChange <= -log2fc_thresh & padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)
# Combine them into one dataframe
highlight_genes <- bind_rows(top_up_P2, top_up_S)
write.csv(highlight_genes, "log2FC_top10genes_P2vsS_2025_06_26.csv", row.names = FALSE)

P2vsS <- ggplot(res_df_P2vsS, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), size = 3.5, alpha = 0.7) +
  scale_color_manual(values = custom_colors3) +
  geom_vline(xintercept = c(-log2fc_thresh, log2fc_thresh), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey80") +
  #geom_text_repel(aes(label = gene), size = 3.5, min.segment.length = 0) +
  labs(x = "Log2FC",
      y = "-log10 adj.p-value",
       color = "") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",  # remove legend if not needed
    panel.grid = element_blank(),
     axis.text = element_text(size = 14),      # increase tick label size
    axis.title = element_text(size = 16),
    axis.line = element_line(color = "black", linewidth = 0.5),
    plot.background = element_rect(fill = "white", color = NA))

# Add tooltip info column
res_df_P2vsS$tooltip <- paste0(
  "Gene: ", res_df_P2vsS$gene, "\n",
  "log2FC: ", round(res_df_P2vsS$log2FoldChange, 2), "\n",
  "-log10(padj): ", round(-log10(res_df_P2vsS$padj), 2)
)
highlight_genes$tooltip <- paste0(
  "Gene: ", highlight_genes$gene, "\n",
  "log2FC: ", round(highlight_genes$log2FoldChange, 2), "\n",
  "-log10(padj): ", round(-log10(highlight_genes$padj), 2)
)
# Create ggplot with 'text' aesthetic
p <- ggplot(res_df_P2vsS, aes(x = log2FoldChange, y = -log10(padj), text = tooltip)) +
  geom_point(aes(color = significance), size = 2, alpha = 0.7) +
  geom_point(data = highlight_genes,
             aes(x = log2FoldChange, y = -log10(padj), text = tooltip),
             shape = 21, stroke = 1, color = "red", fill = NA, size = 3) +
  geom_vline(xintercept = c(-log2fc_thresh, log2fc_thresh), linetype = "dashed", color = "grey80") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey80") +
  labs(x = "Log2FC (P2 vs S)", y = "-log10 adj. p-value", color = "") +
  scale_color_manual(values = custom_colors) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )

P2vsS <- ggplotly(p, tooltip = "text")
saveWidget(P2vsS, file = "volcano_plot_P2_vs_S.html")

```


combine vulcanos - P1vsS, P1vsP2, P2vsS

```{r}
library(cowplot)
combined_vulcano <- plot_grid(P1vsP2, P1vsS, P2vsS, ncol = 3, labels = c("A", "B", "C"),
                              label_size = 16, align = "hv")
ggsave("combined_vulcano.png", combined_vulcano, width = 9, height = 6, dpi = 300)
```



```{r corrected volcano P1 vs P2}
load("transformed_counts_mindepth10.RData")

# ── Load gene names ───────────────────────────────────────────────────────────
gene_names <- read.delim("./geneID2geneNAME.tab", header = TRUE, stringsAsFactors = FALSE)
gene_names$external_gene_name[gene_names$external_gene_name == ""] <- NA

# ── Results ───────────────────────────────────────────────────────────────────
res_P1vsP2    <- results(dds, contrast = c("fish_type2", "P1", "P2"))  #
res_df_P1vsP2 <- as.data.frame(res_P1vsP2) %>%                         # 
  rownames_to_column("ensembl_gene_id") %>%
  left_join(gene_names, by = "ensembl_gene_id") %>%
  mutate(
    external_gene_name = na_if(external_gene_name, ""),
    gene_label = if_else(is.na(external_gene_name), ensembl_gene_id, external_gene_name)
  )

# ── Significance: padj only ───────────────────────────────────────────────────
padj_thresh <- 0.05

res_df_P1vsP2$significance <- "Not Sig"                                
res_df_P1vsP2$significance[                                            
  res_df_P1vsP2$padj < padj_thresh & res_df_P1vsP2$log2FoldChange > 0
] <- "Up in P1"
res_df_P1vsP2$significance[                                            
  res_df_P1vsP2$padj < padj_thresh & res_df_P1vsP2$log2FoldChange < 0
] <- "Up in P2"                                                        

res_df_P1vsP2$significance <- factor(                               
  res_df_P1vsP2$significance,
  levels = c("Up in P1", "Up in P2", "Not Sig")                      
)

custom_colors1 <- c(
  "Up in P1" = "magenta3",       
  "Up in P2" = "purple3",
  "Not Sig" = "black")

# ── Top genes: ranked by padj only ───────────────────────────────────────────
top_up_P1 <- res_df_P1vsP2 %>%                                         
  filter(log2FoldChange > 0, padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

top_up_P2 <- res_df_P1vsP2 %>%                                         
  filter(log2FoldChange < 0, padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

highlight_genes <- bind_rows(top_up_P1, top_up_P2)                    
write.csv(highlight_genes, "../Data/top10genes_P1vsP2_nofcthresh.csv", row.names = FALSE) 

# ── Calculate y-axis max to set label nudge distance ─────────────────────────
y_max <- max(-log10(res_df_P1vsP2$padj), na.rm = TRUE)                

# ── Static volcano ────────────────────────────────────────────────────────────
P1vsP2_static <- ggplot(                                               
  res_df_P1vsP2,                                                       
  aes(x = log2FoldChange, y = -log10(padj))
) +
  geom_point(aes(color = significance), size = 4, alpha = 0.8) +
  scale_color_manual(
    values = custom_colors1,
    guide  = guide_legend(override.aes = list(size = 3, alpha = 1))
  ) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey60") +
  geom_text_repel(
    data               = highlight_genes,
    aes(label          = gene_label),
    size               = 3,
    max.overlaps       = Inf,
    min.segment.length = 0,
    segment.color      = "black",
    segment.size       = 0.5,
    segment.curvature  = 0,
    segment.angle      = 90,
    direction          = "both",
    nudge_y            = y_max * 0.5,
    vjust              = 0,
    force              = 4,
    force_pull         = 0.01,
    box.padding        = 1,             # ← more space around each label
    point.padding      = 0.5           # ← more space between label and point
  ) +
  labs(
    x     = expression(log[2]~"Fold Change"),
    y     = expression(-log[10]~"adj. p-value"),
    color = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = c(0.15, 0.1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.text       = element_text(size = 9),
    panel.grid        = element_blank(),
    axis.text         = element_text(size = 14),
    axis.title        = element_text(size = 16),
    axis.line         = element_line(color = "black", linewidth = 0.5),
    plot.background   = element_rect(fill = "white", color = NA)
  )

P1vsP2_static
```


```{r corrected volcano plot P1 vs S}
load("transformed_counts_mindepth10.RData")

# ── Load gene names ───────────────────────────────────────────────────────────
gene_names <- read.delim("./geneID2geneNAME.tab", header = TRUE, stringsAsFactors = FALSE)
gene_names$external_gene_name[gene_names$external_gene_name == ""] <- NA

# ── Results ───────────────────────────────────────────────────────────────────
res_P1vsS    <- results(dds, contrast = c("fish_type2", "P1", "S"))
res_df_P1vsS <- as.data.frame(res_P1vsS) %>%
  rownames_to_column("ensembl_gene_id") %>%
  left_join(gene_names, by = "ensembl_gene_id") %>%
  mutate(
    external_gene_name = na_if(external_gene_name, ""),
    gene_label = if_else(is.na(external_gene_name), ensembl_gene_id, external_gene_name)
  )

# ── Significance: padj only ───────────────────────────────────────────────────
padj_thresh <- 0.05

res_df_P1vsS$significance <- "Not Sig"
res_df_P1vsS$significance[
  res_df_P1vsS$padj < padj_thresh & res_df_P1vsS$log2FoldChange > 0
] <- "Up in P1"
res_df_P1vsS$significance[
  res_df_P1vsS$padj < padj_thresh & res_df_P1vsS$log2FoldChange < 0
] <- "Up in S"

res_df_P1vsS$significance <- factor(
  res_df_P1vsS$significance,
  levels = c("Up in P1", "Up in S", "Not Sig")
)

custom_colors <- c(
  "Up in P1" = "magenta3",
  "Up in S"  = "gray40",   # purple to match reference image
  "Not Sig"  = "black"
)


# ── Top genes: ranked by padj only ───────────────────────────────────────────
top_up_P1 <- res_df_P1vsS %>%
  filter(log2FoldChange > 0, padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

top_up_S <- res_df_P1vsS %>%
  filter(log2FoldChange < 0, padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

highlight_genes <- bind_rows(top_up_P1, top_up_S)
write.csv(highlight_genes, "../Data/top10genes_P1vsS_nofcthresh.csv", row.names = FALSE)

# ── Calculate y-axis max to set label nudge distance ─────────────────────────
y_max <- max(-log10(res_df_P1vsS$padj), na.rm = TRUE)

# ── Static volcano ────────────────────────────────────────────────────────────
P1vsS_static <- ggplot(
  res_df_P1vsS,
  aes(x = log2FoldChange, y = -log10(padj))
) +
  geom_point(aes(color = significance), size = 4, alpha = 0.8) +
  scale_color_manual(
    values = custom_colors,
    guide  = guide_legend(override.aes = list(size = 3, alpha = 1))  # larger legend dots
  ) +
  # ── visual reference lines (not used for DEG classification) ────────────────
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey60") +

  # ── gene labels: straight vertical lines going up to labels ──────────────────
  geom_text_repel(
   data               = highlight_genes,
    aes(label          = gene_label),
    size               = 3,
    max.overlaps       = Inf,
    min.segment.length = 0,
    segment.color      = "black",
    segment.size       = 0.5,
    segment.curvature  = 0,
    segment.angle      = 90,
    direction          = "both",
    nudge_y            = y_max * 0.5,
    vjust              = 0,
    force              = 4,
    force_pull         = 0.01,
    box.padding        = 1,             # ← more space around each label
    point.padding      = 0.5           # ← more space between label and point
  ) +
  labs(
    x     = expression(log[2]~"Fold Change"),
    y     = expression(-log[10]~"adj. p-value"),
    color = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = c(0.15, 0.1),   # ← bottom left like reference
    legend.background = element_rect(fill = "white", color = NA),
    legend.text       = element_text(size = 9),
    panel.grid        = element_blank(),
    axis.text         = element_text(size = 14),
    axis.title        = element_text(size = 16),
    axis.line         = element_line(color = "black", linewidth = 0.5),
    plot.background   = element_rect(fill = "white", color = NA)
  )

P1vsS_static
```


```{r corrected volcano P2 vs S}
load("transformed_counts_mindepth10.RData")

# ── Load gene names ───────────────────────────────────────────────────────────
gene_names <- read.delim("./geneID2geneNAME.tab", header = TRUE, stringsAsFactors = FALSE)
gene_names$external_gene_name[gene_names$external_gene_name == ""] <- NA

# ── Results ───────────────────────────────────────────────────────────────────
res_P2vsS    <- results(dds, contrast = c("fish_type2", "P2", "S"))    
res_df_P2vsS <- as.data.frame(res_P2vsS) %>%                           
  rownames_to_column("ensembl_gene_id") %>%
  left_join(gene_names, by = "ensembl_gene_id") %>%
  mutate(
    external_gene_name = na_if(external_gene_name, ""),
    gene_label = if_else(is.na(external_gene_name), ensembl_gene_id, external_gene_name)
  )

# ── Significance: padj only ───────────────────────────────────────────────────
padj_thresh <- 0.05

res_df_P2vsS$significance <- "Not Sig"                                 
res_df_P2vsS$significance[                                             
  res_df_P2vsS$padj < padj_thresh & res_df_P2vsS$log2FoldChange > 0
] <- "Up in P2"                                                        
res_df_P2vsS$significance[                                             
  res_df_P2vsS$padj < padj_thresh & res_df_P2vsS$log2FoldChange < 0
] <- "Up in S"                                                         

res_df_P2vsS$significance <- factor(                                   
  res_df_P2vsS$significance,
  levels = c("Up in P2", "Up in S", "Not Sig")                        
)

custom_colors2 <- c(                                                   
  "Up in P2" = "purple3",                                              
  "Up in S"  = "gray40",                                               
  "Not Sig"  = "black"
)

# ── Top genes: ranked by padj only ───────────────────────────────────────────
top_up_P2 <- res_df_P2vsS %>%                                          
  filter(log2FoldChange > 0, padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

top_up_S <- res_df_P2vsS %>%                                          
  filter(log2FoldChange < 0, padj < padj_thresh) %>%
  arrange(padj) %>%
  slice_head(n = 10)

highlight_genes <- bind_rows(top_up_P2, top_up_S)                     
write.csv(highlight_genes, "../Data/top10genes_P2vsS_nofcthresh.csv", row.names = FALSE)  

# ── Calculate y-axis max to set label nudge distance ─────────────────────────
y_max <- max(-log10(res_df_P2vsS$padj), na.rm = TRUE)                  

# ── Static volcano ────────────────────────────────────────────────────────────
P2vsS_static <- ggplot(                                               
  res_df_P2vsS,                                                        
  aes(x = log2FoldChange, y = -log10(padj))
) +
  geom_point(aes(color = significance), size = 4, alpha = 0.8) +
  scale_color_manual(
    values = custom_colors2,                                           
    guide  = guide_legend(override.aes = list(size = 3, alpha = 1))
  ) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(padj_thresh), linetype = "dashed", color = "grey60") +
  geom_text_repel(
    data               = highlight_genes,
    aes(label          = gene_label),
    size               = 3.5,
    max.overlaps       = Inf,
    min.segment.length = 0,
    segment.color      = "black",
    segment.size       = 0.5,
    segment.curvature  = 0,
    segment.angle      = 90,
    direction          = "x",
    nudge_y            = y_max * 0.3,
    vjust              = 0,
    force              = 3,
    force_pull         = 0.03
  ) +
  labs(
    x     = expression(log[2]~"Fold Change"),
    y     = expression(-log[10]~"adj. p-value"),
    color = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = c(0.25, 0.8),
    legend.background = element_rect(fill = "white", color = NA),
    legend.text       = element_text(size = 9),
    panel.grid        = element_blank(),
    axis.text         = element_text(size = 14),
    axis.title        = element_text(size = 16),
    axis.line         = element_line(color = "black", linewidth = 0.5),
    plot.background   = element_rect(fill = "white", color = NA)
  )

P2vsS_static                                                           
```



```{r corrected vulcano combined}
library(cowplot)
combined_vulcano <- plot_grid(P1vsP2_static, P1vsS_static, ncol = 2, labels = c("A", "B"),
                              label_size = 16, align = "hv")
ggsave("../Figures/combined_vulcano_top10named_2only20260407.png", combined_vulcano, width = 10, height = 7, dpi = 1200)
ggsave("../Figures/combined_vulcano_top10named_2only20260407.pdf", combined_vulcano, width = 10, height = 7, dpi = 1200)

p2vss<- plot_grid(P2vsS_static, ncol = 1, labels = c("C"),
                              label_size = 16, align = "hv")
#save P2vsS separately and add as a small insert
ggsave("../Figures/P2vsS_20260407.png", p2vss, width = 3, height = 5, dpi = 1200)



# Create inset plot positioned inside P1vsS
inset_plot <- ggdraw(P1vsS_static) +
  draw_plot(P1vsS_static, x = 0.8, y = 0.7, width = 0.3, height = 0.3) +  # adjust position/size
  draw_plot_label(label = "C", x = 0.8, y = 0.95, size = 14)        # label for the inset

# Combine with P1vsP2
combined_vulcano_insert <- plot_grid(P1vsP2, inset_plot, labels = c("A", "B"), label_size = 16)
```